---
title: "Exploratory Data Analysis"
author: "David Sprague"
date: "2024-01-20"
output:
  html_document:
    theme: paper
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Health Insurance Rates in the USA

### Introduction 

In this exploratory data analysis, I look at the percentage of people aged 19-34 in each state in the United States that have at least one form of health insurance. The objective of this study is to closely look at the geographic patterns and distribution of people with health insurance. Health care in the united states can be far more expensive than most people can reasonably afford. While health care is not always affordable, it is a way to reduce costs for health care. Many full-time jobs offer health insurance options but for those that do not and for people who are unemployed, it can be costly to buy health insurance. 


### Data Preparation 
```{r libraries, message = FALSE, warning = FALSE}
# Load libraries
library(sf)
library(tidyverse)
library(spdep)
library(kableExtra)
library(tmap)
```




```{r data_prep, message = FALSE, warning = FALSE}
## Read in 2020 health insurance data 
insurance_data <- read_csv(file = "../raw_data/usa_health_insurance_data.csv")


## Read in Spatial data layer
states <- read_sf("../raw_data/usa_states.gpkg")


## Select columns needed from insurance data
insurance_data <- insurance_data |> select(c(GEOID, 
                                            B27010_018E, 
                                            B27010_019E))
## Rename columns in insurance data table 
insurance_data <- insurance_data |> rename(population_19_34 = B27010_018E,
                                           people_with_health_insurance = B27010_019E)

## Create column for percentage of people who have health insurance in each state aged 19-34
insurance_data <- insurance_data |> mutate(percent_has_insurance = people_with_health_insurance / population_19_34 * 100)


## Select only GEOID,  GEOM, and NAME columns from spatial data layer
states <- states |> select(GEOID, NAME, geom)


## Join spatial data layer with insurance data
joined_insurance_spatial_data <- left_join(states, insurance_data, by = "GEOID")

## Remove any rows containing NA values
joined_insurance_spatial_data <- joined_insurance_spatial_data |> na.omit()
```


For this data analysis, I sourced my health insurance data from the American Community Survey API. The spatial data layer for the United States was acquired through the 'Tigris' library. Processing the data involved selecting necessary columns from the ACS which included the total number of people aged 19-34 with health insurance in each State. To improve readability of the table, I gave the columns meaningful names. A new column was created to represent the percentage of individuals who have health insurance in each state(19-34 years old). The last step of wrangling the data involved data alignment, ensuring that the health insurance and spatial data tables would properly join using a left table join. 

***

### Exploratory Spatial Data Analysis


#### Data Summary
```{r data_summary, message = FALSE, warning = FALSE}
## Calculate number of rows 
num_obs <- joined_insurance_spatial_data |> nrow()


## Calculate the minimum percent for people with insurance in each state
min_percent <- min(joined_insurance_spatial_data$percent_has_insurance) |> round(2)


## Calculate the maximum percent for people with insurance in each state
max_percent <- max(joined_insurance_spatial_data$percent_has_insurance) |> round(2)


## Calculate the mean percentage for people with insurance in each state
mean_percent <- mean(joined_insurance_spatial_data$percent_has_insurance) |> round(2)


## Calculate standard deviation for 2020
standard_deviation <- sd(joined_insurance_spatial_data$percent_has_insurance) |> round(2)


## Create histogram for percentage of people with insurance in each state
## Create histogram for 2020
joined_insurance_spatial_data |> 
  ggplot(mapping = aes(x = joined_insurance_spatial_data$percent_has_insurance)) +
  geom_histogram(binwidth = 1,
                 color = "black",
                 fill = "lightblue") +
  labs(title = "United States Health Insurance",
       y = "Number of States",
       x = "% of people that have at least one form of health insurance") + 
  theme_classic()
```

The variable of interst in this data analysis is the percentage of people aged 19-34 in each state in the United States that has at least one form of health insurance in the year 2020. The data set had `r num_obs` observations. In 2020, the minimum, maximum, and mean percentage of people who had health insurance was `r min_percent`, `r max_percent`, and `r mean_percent` The standard deviation was `r standard_deviation` From the histogram, we can see that the data is fairly left skewed. 


***

#### Geographic Distribution and Clustering
```{r map, message = FALSE, warning = FALSE}
## Create Queen case neighbors
US_queen_neighbors <- poly2nb(joined_insurance_spatial_data,
                    queen = TRUE)


## Convert neighbors to weight matrix 
US_queen_weights <- nb2listw(US_queen_neighbors,
                         style = "B",         
                         zero.policy = TRUE)


## Moran's I Analysis
US_insurance_morans <- moran.test(joined_insurance_spatial_data$percent_has_insurance,      
                             US_queen_weights,          
                             zero.policy = TRUE,    
                             randomisation = TRUE)  


## Store Moran's I value for percentage of people who have insurance 
US_morans <- US_insurance_morans$estimate[1] |> as.double() |> round(4)


## Calculate LISA statistic for percentage of people who have insurance
US_lisa_statistic <- localmoran(joined_insurance_spatial_data$percent_has_insurance,
                          listw = US_queen_weights,      
                          alternative = "two.sided", 
                          zero.policy = TRUE) |>     
  as_tibble() |>                                     
  mutate(across(everything(), as.vector))


## Add values required for LISA category
US_lisa_statistic <- US_lisa_statistic |>
  mutate(SCVAR =  scale(joined_insurance_spatial_data$percent_has_insurance) |> as.vector(),
         LAGVAR = lag.listw(US_queen_weights, scale(joined_insurance_spatial_data$percent_has_insurance)),
         LISACAT = case_when(SCVAR >= 0 & LAGVAR >= 0 & `Pr(z != E(Ii))` <= 0.05 ~ 1,
                             SCVAR <= 0 & LAGVAR <= 0 & `Pr(z != E(Ii))` <= 0.05 ~ 2,
                             SCVAR >= 0 & LAGVAR <= 0 & `Pr(z != E(Ii))` <= 0.05 ~ 3,
                             SCVAR <= 0 & LAGVAR >= 0 & `Pr(z != E(Ii))` <= 0.05 ~ 4,
                             `Pr(z != E(Ii))` > 0.05 ~ 5))


## Add label based on the values
US_lisa_statistic <- US_lisa_statistic |>
  mutate(CATNAME = case_when(LISACAT == 1 ~ "High-High",
                             LISACAT == 2 ~ "Low-Low",
                             LISACAT == 3 ~ "High-Low",
                             LISACAT == 4 ~ "Low-High",
                             LISACAT == 5 ~ "Not Significant"))


## Add LISA category column to the spatial data for mapping
joined_insurance_spatial_data <- joined_insurance_spatial_data |>
  mutate(LISACAT = US_lisa_statistic$LISACAT,
         CATNAME = US_lisa_statistic$CATNAME)


## Count for LISA high-high states
hh_count <- table(US_lisa_statistic$CATNAME)["High-High"] |> as.integer()



## Count for LISA low-low states 
lh_count <- table(US_lisa_statistic$CATNAME)["Low-Low"] |> as.integer()


## Create kable table to summarize lisa results and add header above
US_lisa_statistic$CATNAME |> 
      table() |> 
      kable(col.names = c("Outliers", "Frequency"), 
            align = "l") |>
      kable_styling(bootstrap_options = c("striped", 
                                          "hover", 
                                          "condensed", 
                                          "responsive"),
                    full_width = T) |> add_header_above(c("Lisa Results" = 2), align = "left")


## Adjust tmap mode to be "view" 
tmap_mode("view")

## Change na values in LISACAT column to be 5(Not significant)
joined_insurance_spatial_data$LISACAT[is.na(joined_insurance_spatial_data$LISACAT)] <- 5



## Make choropleth map for percentage of people with health insurance 
health_insurance_map <- joined_insurance_spatial_data |>
  tm_shape() +
    tm_polygons(col = "percent_has_insurance",
              title = "People with health insurance (%)", 
              style = "jenks",
              colorNA = "white",
              alpha = 0.9,
              border.col = "black",
              border.alpha = 0.3,
              id = "NAME",
              palette = "Greens") +
    tm_layout(frame = FALSE, 
              main.title = "Health insurance in the US",
              main.title.size = 1.7, 
              legend.outside = TRUE) 


## Construct the LISA map
lisa_map <- tm_shape(joined_insurance_spatial_data) +
  tm_polygons(col = "grey50") +
tm_shape(joined_insurance_spatial_data) + 
  tm_polygons("LISACAT", 
              title = "LISA Category",
              breaks = c(1, 2, 3, 4, 5, 6),
              palette =  c("red", 
                           "blue", 
                           "lightpink", 
                           "skyblue", 
                           "grey90"),
              colorNA = "white",
              labels = c("High-High", 
                         "Low-Low",
                         "High-Low",
                         "Low-High", 
                         "Not significant"),
              border.col = "black", 
              id = "NAME",
              border.alpha = 0.25) +
  tm_layout(frame = FALSE,
            legend.outside = TRUE)


## Map the two maps synced in view mode 
tmap_arrange(health_insurance_map, 
             lisa_map,
             ncol = 2,
             nrow = 1, 
             sync = TRUE)
```


### Conclusions



